---
title: "EDA_Abhimanyu"
author: "Carrie Magee, Abhimanyu Barun, Ambar Pathak, Kismat Khatri"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ezids)
library(ggplot2)
library(dplyr)
library(corrplot)

dataset = read.csv("sadc_df.csv")
# select variables v1, v2, v3
myvars <- c("Age_Alc", "Age_Weed","multiple_partners","physical_activity","Hrs_Sleep","age_sex","race","gender","fight","suicide")
data <- dataset[myvars]
data <- data[data$age_sex != 0,]

data$race <- as.factor(data$race)
data$gender <- as.factor(data$gender)
data$fight <- as.factor(data$fight)
data$suicide <- as.factor(data$suicide)

data$fight <- recode(data$fight, '0' = 1, '1' = 0)
data$suicide <- recode(data$suicide, '0' = 1, '1' = 0)

```

## R Markdown


```{r}

# Age when they started drinking

data_age_alc <- data %>% 
              group_by(Age_Alc) %>% 
              summarise(total_count = n(), .groups = "drop") %>% 
              as.data.frame()

str(data_age_alc)

# Plot the bar chart 
ggplot(data_age_alc, aes(x = Age_Alc, y = total_count)) + 
  geom_bar(stat = "identity", fill="forest green") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Engaging in a Alcoholism by Age when Individual Started Drinking") +
  xlab("Age") +
  ylab("Count")

```

This bar graph illustrates the count of respondents who had their first drink for age from 8 to 13 <correct english abit>. We observe that most respondents started drinking when they were 10 and 12. There are around 500 people who started drink at age 8 and the number increases with 500 as the age increases to 11. Overall we can see that there is an upward trend till age 11 followed by a decline till age 13. 


```{r}

# Age when they started smoking

data_age_weed <- data %>% 
              group_by(Age_Weed) %>% 
              summarise(total_count = n(), .groups = 'drop') %>% 
              as.data.frame()

str(data_age_weed)

# Plot the bar chart 
ggplot(data_age_weed, aes(x = Age_Weed, y = total_count)) + 
  geom_bar(stat = "identity", fill="forest green") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Engaging in Smoking Weed by Age when Individual Started Smoking") +
  xlab("Age") +
  ylab("Count")

```

This bar graph illustrates the count of respondents who smoked weed for the fisr time for age from 8 to 13 <correct english abit>. We observe that few respondents started smoking weed as early as at 8 years. There are more people who starter smokign ar age 9 (almost 1500) than at age 11 (almost 1000). There are around 1750 respondents whi had their first smoke at age 11 and 12.  


```{r}
  
# Age when the had sex
data_age_sex <- data %>% 
              group_by(age_sex) %>% 
              summarise(total_count = n(), .groups = 'drop') %>% 
              as.data.frame()

str(data_age_sex)

# Plot the bar chart 
ggplot(data_age_sex, aes(x = age_sex, y = total_count)) + 
  geom_bar(stat = "identity", fill="forest green") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Engaging in a Intercourse by Age when Individual Started having Intercourse") +
  xlab("Age") +
  ylab("Count")

```

From the graph above we see that most respondents had their first sexual experience at ages 10, 11 and 13. There are barely 750 respondets whi had their first experience before the age of 10. There is a huge increase in numbers when we go from age 9 to age 10.

```{r}

# Multiple partners
data_age_partner <- data %>% 
              group_by(multiple_partners) %>% 
              summarise(total_count = n(), .groups = 'drop') %>% 
              as.data.frame()

str(data_age_partner)

# Plot the bar chart 
ggplot(data_age_partner, aes(x = multiple_partners, y = total_count)) + 
  geom_bar(stat = "identity", fill="forest green") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Having Multiple Partners by Age") +
  xlab("Age") +
  ylab("Count")

```

We observe that most respondents have 1 partner followed by 6, 3, 2, 1 and 4. Majority of the respondents have either 1 or 6 partners. 

```{r}

# Physical Activity
data_physical_activity <- data %>% 
              group_by(physical_activity) %>% 
              summarise(total_count = n(), .groups = 'drop') %>% 
              as.data.frame()

str(data_physical_activity)

# Plot the bar chart 
ggplot(data_physical_activity, aes(x = physical_activity, y = total_count)) + 
  geom_bar(stat = "identity", fill="forest green") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("No. of Days a Week the respondent Exercise by Age") +
  xlab("Age") +
  ylab("Count")

```

Most respondents(over 4000) exercise 7 days a week.

```{r}
  
data_grouped_2a <- data %>% 
              group_by(gender, multiple_partners) %>% 
              summarise(total_count1 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_2a, aes(fill=multiple_partners, y=total_count1, x=multiple_partners)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~gender) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Multiple Partners of Respondents of Age for different genders") +
    xlab("Age") +
    ylab("Count")

```
On observing the bar plot for multiple partners for both genders, we can say that overall males have more partners than females. There are more number of males who have 6 partners. But there are more number of females whi have 5 partners. There is a huge gap in most of the bar graphs  because there are 3 times more males than females.

```{r}

data_grouped_2b <- data %>% 
              group_by(gender, Age_Weed) %>% 
              summarise(total_count2 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_2b, aes(fill=Age_Weed, y=total_count2, x=Age_Weed)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~gender) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Age whe respondent Smoked Weed for different genders by Age") +
    xlab("Age") +
    ylab("Count")

```

Most males first smoked weed at the age of 8, 11, 12 and 13 while most females first smoked weed at 10, 11 and 12. 

```{r}
  
data_grouped_2c <- data %>% 
              group_by(gender, Age_Alc) %>% 
              summarise(total_count3 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_2c, aes(fill=Age_Alc, y=total_count3, x=Age_Alc)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~gender) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Age when respondent had first Alcohol Drink for different genders") +
    xlab("Age") +
    ylab("Count")

```

Most males had their first alcohol at ages 10, 11, 12 and 13. There is an increasing trend till age 11 and a declining trend after that. For females most had their first drink at ages 9, 11 and 12. 

```{r}

data_grouped_3a <- data %>% 
              group_by(race, multiple_partners) %>% 
              summarise(total_count1 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_3a, aes(fill=multiple_partners, y=total_count1, x=multiple_partners)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Multiple partners of Respondents by Age for different Ethnic Groups") +
    xlab("Age") +
    ylab("Count")

```
On comparing multiple partners for different ethnic groups, we observe that majorly there is a decreasing trend. Number of people with multiple partners decrease as the number of partners increase. For ethnic group 2 we observe that most people have 6 multiple partners while similar number of people have 1 to 5 partners. Most people of ethnic group 3 have 1, 3 or 5 partners as opposed. Similar to ethnic group 1, group 3 also has a decreasing trend. Number of people with multiple partners decrease as the number of partners increase.

```{r}

data_grouped_3b <- data %>% 
              group_by(race, Age_Weed) %>% 
              summarise(total_count2 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_3b, aes(fill=Age_Weed, y=total_count2, x=Age_Weed)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Age when respondent Smoked Weed the first time for different Ethnic Groups") +
    xlab("Age") +
    ylab("Count")

```
Most respondents belonging to ethnic group 1 smoked weed for the first time at age 11 and 12. There is an increasing trend form 8 to 12. For ethnic group 2, we observe that most experienced weed at age 9. For ethnic group 3, most of the respondents were atleast 10 years old. Most Respondents from ethnic group 4 were also atleast 10 years old when they smoked weed for the first time. 

```{r}
  
data_grouped_3c <- data %>% 
              group_by(race, Age_Alc) %>% 
              summarise(total_count3 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_3c, aes(fill=Age_Alc, y=total_count3, x=Age_Alc)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Age when respondent had first Alcohol Drink for different Ethnic Groups") +
    xlab("Age") +
    ylab("Count")

```

Most Respondents from Ethnic group 1 had alcohol for the first time when they were atleast 11 years old. Less than 500 of them had alcohol before the age of 11. At age 10, the respondents from ethnic group 2 had alcohol. For Ethnic group 3, most of them had alcohol at age of atleast 11 years. There are few (around 560) who had alcohol at age 9. For group 4, there is an increasing trend ie as the age increases respondents who have experienced alcohol also increased.
 
```{r}

data_grouped_4a <- data %>% 
              group_by(gender, suicide) %>% 
              summarise(total_count1 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_4a, aes(fill=suicide, y=total_count1, x=suicide)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~gender) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Age when respondent had Suicidal thoughts for different Genders") +
    xlab("Age") +
    ylab("Count")

```

Overall more males have had suicidal thoughts than females. Around 40% of the males have had suicidal thoughts while only 6% of the females have had suicidal thoughts. 

```{r}

data_grouped_4b <- data %>% 
              group_by(race, suicide) %>% 
              summarise(total_count2 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_4b, aes(fill=suicide, y=total_count2, x=suicide)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Age when respondent had Suicidal thoughts for different Ethnic Groups") +
    xlab("Age") +
    ylab("Count")

```

About 10 Percent (1 in 10 people) of respondents belonging to  ethnic group 1 have had suicidal thoughts. This is not the case with respondents belonging to ethnic group 2. Most of the respondents(about 75%) have had suicidal thoughts. We can say that every 5th person ethnic groups 3 and 4 have had suicidal thoughts.


```{r}

data_grouped_5a <- data %>% 
              group_by(gender, fight) %>% 
              summarise(total_count1 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_5a, aes(fill=fight, y=total_count1, x=fight)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~gender) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Age when respondent had Fights for different Genders") +
    xlab("Age") +
    ylab("Count")

```

Most of the males and females haev been in a fight. There is a higher percentage of females who have been in a fight than men. 


```{r}

data_grouped_5b <- data %>% 
              group_by(race, fight) %>% 
              summarise(total_count2 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_5b, aes(fill=fight, y=total_count2, x=fight)) + 
    geom_bar(position="dodge", stat="identity", fill="forest green") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(plot.title = element_text(hjust = 0.5)) +
    ggtitle("Observing Age when respondent had Fights for different Ethnic Groups") +
    xlab("Age") +
    ylab("Count")

```

Most respondents belonging to ethnic group 1 and 4 have not been in fights while most respondents from ethnic group 2 and 3have been in a fight. 
