---
title: "EDA"
output: html_document
date: "2022-12-05"
---

Adolescence is a crucial life period marked by many different physical, emotional social, academic, and interpersonal changes. The attitudes and  behaviors that are developed in adolescence tend to influence the trajectory of our lives into young adulthood and beyond which makes it such a compelling area of research. With data collected from the CDC’s Youth Risk Behavior Surveillance System (YRBSS), our project explores the questions “Does the age at which adolescents start engaging in "risky" behaviors predict their interpersonal, physical, and mental well-being? Furthermore, how does race influence interpersonal, physical, and mental well-being?”

We conceptualized and measured risky behaviors using the age when individuals started drinking alcohol, smoking marijuana, and having romantic relationships. Considering the objective of the project to explore risky behaviors in early adolescence, we discarded any responses that reported never engaging in any risky behaviors. The resulting ages ranged between 8 years and 13 years old.

Interpersonal, physical, and mental well-being were explored using amount of  physical activity, hours of sleep per night, number of sexual partners, thoughts of suicide, and whether or not individuals have ever engaged in a fight with another person. More specifically, physical activity was measured with the question “During the past 7 days, on how many days were you physically active for a total of at least 60 minutes per day?” with responses ranging from 0 to 7 days. Sexual partners was measured by asking “With how many people have you ever had sexual intercourse?” with answers ranging from 1 to 6 people. For hours of the sleep, the question stated “On an average school night, how many hours of sleep do you get?” and answers ranged from 4 hours to 10 hours. Lastly, suicidal thoughts and fighting were binary coded (yes/no) and asked “Have you ever seriously thought about killing yourself?” and “Have you ever been in a physical fight?” 

Through our analysis, we hope to understand how engaging in risky behaviors either earlier or later in adolescence may predict future interpersonal, physical, and mental well-being for youth. We hope to expand understanding about how risky behaviors can impact the way in which youth function in the various realms of everyday life and uncover any patterns that can lead to future prevent efforts for adolescents. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE )
library(ezids)
library(ggplot2)
library(dplyr)
library(corrplot) 
library(data.table)
library(formattable)
library(tidyr)
```


```{r}
dataset = read.csv("sadc_df.csv")
myvars <- c("Age_Alc", "Age_Weed","multiple_partners","physical_activity","Hrs_Sleep","age_sex","race","sex","fight","suicide")
data <- dataset[myvars]
data <- data[data$age_sex != 0,] 
data$race <- as.factor(data$race)
```

```{r}
data$fight <- recode(data$fight, '0' = 'yes', '1' = 'no')
data$suicide <- recode(data$suicide, '0' = 'yes', '1' = 'no')

no_suicide <- subset(data,suicide == 'no')
yes_suicide <- subset(data,suicide == 'yes')

no_fight<- subset(data,fight == 'no')
yes_fight <- subset(data,fight == 'yes')
```


```{r}
# Age when they started drinking

data_age_alc <- data %>% 
              group_by(Age_Alc) %>% 
              summarise(total_count = n(), .groups = "drop") %>% 
              as.data.frame()

str(data_age_alc)

# Plot the bar chart 
ggplot(data_age_alc, aes(x = Age_Alc, y = total_count)) + 
  geom_bar(stat = "identity")

```

This bar graph illustrates the count of respondents who had their first drink for age from 8 to 13 <correct english abit>. We observe that most respondents started drinking when they were 10 and 12. There are around 500 people who started drink at age 8 and the number increases with 500 as the age increases to 11. Overall we can see that there is an upward trend till age 11 followed by a decline till age 13. 


```{r}
# Age when they started smoking

data_age_weed <- data %>% 
              group_by(Age_Weed) %>% 
              summarise(total_count = n(), .groups = 'drop') %>% 
              as.data.frame()

str(data_age_weed)

# Plot the bar chart 
ggplot(data_age_weed, aes(x = Age_Weed, y = total_count)) + 
  geom_bar(stat = "identity")

```

This bar graph illustrates the count of respondents who smoked weed for the fisr time for age from 8 to 13 <correct english abit>. We observe that few respondents started smoking weed as early as at 8 years. There are more people who starter smokign ar age 9 (almost 1500) than at age 11 (almost 1000). There are around 1750 respondents whi had their first smoke at age 11 and 12.  


```{r}
# Age when the had sex
data_age_sex <- data %>% 
              group_by(age_sex) %>% 
              summarise(total_count = n(), .groups = 'drop') %>% 
              as.data.frame()

str(data_age_sex)

# Plot the bar chart 
ggplot(data_age_sex, aes(x = age_sex, y = total_count)) + 
  geom_bar(stat = "identity")

```

From the graph above we see that most respondents had their first sexual experience at ages 10, 11 and 13. There are barely 750 respondets whi had their first experience before the age of 10. There is a huge increase in numbers when we go from age 9 to age 10.

```{r}
# Multiple partners
data_age_partner <- data %>% 
              group_by(multiple_partners) %>% 
              summarise(total_count = n(), .groups = 'drop') %>% 
              as.data.frame()

str(data_age_partner)

# Plot the bar chart 
ggplot(data_age_partner, aes(x = multiple_partners, y = total_count)) + 
  geom_bar(stat = "identity")

```

We observe that most respondents have 1 partner followed by 6, 3, 2, 1 and 4. Majority of the respondents have either 1 or 6 partners. 

```{r}
# Physical Activity
data_physical_activity <- data %>% 
              group_by(physical_activity) %>% 
              summarise(total_count = n(), .groups = 'drop') %>% 
              as.data.frame()

str(data_physical_activity)

# Plot the bar chart 
ggplot(data_physical_activity, aes(x = physical_activity, y = total_count)) + 
  geom_bar(stat = "identity")

```

Most respondents(over 4000) exercise 7 days a week.

```{r}
data_grouped_2a <- data %>% 
              group_by(sex, multiple_partners) %>% 
              summarise(total_count1 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_2a, aes(fill=multiple_partners, y=total_count1, x=multiple_partners)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~sex) +
    theme(legend.position="none") +
    xlab("")

```
On observing the bar plot for multiple partners for both genders, we can say that overall males have more partners than females. There are more number of males who have 6 partners. But there are more number of females whi have 5 partners. There is a huge gap in most of the bar graphs  because there are 3 times more males than females.

```{r}
data_grouped_2b <- data %>% 
              group_by(sex, Age_Weed) %>% 
              summarise(total_count2 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_2b, aes(fill=Age_Weed, y=total_count2, x=Age_Weed)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~sex) +
    theme(legend.position="none") +
    xlab("")

```

Most males first smoked weed at the age of 8, 11, 12 and 13 while most females first smoked weed at 10, 11 and 12. 

```{r}
data_grouped_2c <- data %>% 
              group_by(sex, Age_Alc) %>% 
              summarise(total_count3 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_2c, aes(fill=Age_Alc, y=total_count3, x=Age_Alc)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~sex) +
    theme(legend.position="none") +
    xlab("")

```

Most males had their first alcohol at ages 10, 11, 12 and 13. There is an increasing trend till age 11 and a declining trend after that. For females most had their first drink at ages 9, 11 and 12. 

```{r}

data_grouped_3a <- data %>% 
              group_by(race, multiple_partners) %>% 
              summarise(total_count1 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_3a, aes(fill=multiple_partners, y=total_count1, x=multiple_partners)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(legend.position="none") +
    xlab("")

```
On comparing multiple partners for different ethnic groups, we observe that majorly there is a decreasing trend. Number of people with multiple partners decrease as the number of partners increase. For ethnic group 2 we observe that most people have 6 multiple partners while similar number of people have 1 to 5 partners. Most people of ethnic group 3 have 1, 3 or 5 partners as opposed. Similar to ethnic group 1, group 3 also has a decreasing trend. Number of people with multiple partners decrease as the number of partners increase.

```{r}

data_grouped_3b <- data %>% 
              group_by(race, Age_Weed) %>% 
              summarise(total_count2 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_3b, aes(fill=Age_Weed, y=total_count2, x=Age_Weed)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(legend.position="none") +
    xlab("")

```
Most respondents belonging to ethnic group 1 smoked weed for the first time at age 11 and 12. There is an increasing trend form 8 to 12. For ethnic group 2, we observe that most experienced weed at age 9. For ethnic group 3, most of the respondents were atleast 10 years old. Most Respondents from ethnic group 4 were also atleast 10 years old when they smoked weed for the first time. 

```{r}
data_grouped_3c <- data %>% 
              group_by(race, Age_Alc) %>% 
              summarise(total_count3 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_3c, aes(fill=Age_Alc, y=total_count3, x=Age_Alc)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(legend.position="none") +
    xlab("")

```

Most Respondents from Ethnic group 1 had alcohol for the first time when they were atleast 11 years old. Less than 500 of them had alcohol before the age of 11. At age 10, the respondents from ethnic group 2 had alcohol. For Ethnic group 3, most of them had alcohol at age of atleast 11 years. There are few (around 560) who had alcohol at age 9. For group 4, there is an increasing trend ie as the age increases respondents who have experienced alcohol also increased.
 
```{r}
data_grouped_4a <- data %>% 
              group_by(sex, suicide) %>% 
              summarise(total_count1 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_4a, aes(fill=suicide, y=total_count1, x=suicide)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~sex) +
    theme(legend.position="none") +
    xlab("")

```

Overall more males have had suicidal thoughts than females. Around 40% of the males have had suicidal thoughts while only 6% of the females have had suicidal thoughts. 

```{r}
data_grouped_4b <- data %>% 
              group_by(race, suicide) %>% 
              summarise(total_count2 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_4b, aes(fill=suicide, y=total_count2, x=suicide)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(legend.position="none") +
    xlab("")

```

About 10 Percent (1 in 10 people) of respondents belonging to  ethnic group 1 have had suicidal thoughts. This is not the case with respondents belonging to ethnic group 2. Most of the respondents(about 75%) have had suicidal thoughts. We can say that every 5th person ethnic groups 3 and 4 have had suicidal thoughts.


```{r}
data_grouped_5a <- data %>% 
              group_by(sex, fight) %>% 
              summarise(total_count1 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_5a, aes(fill=fight, y=total_count1, x=fight)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~sex) +
    theme(legend.position="none") +
    xlab("")

```

Most of the males and females haev been in a fight. There is a higher percentage of females who have been in a fight than men. 


```{r}

data_grouped_5b <- data %>% 
              group_by(race, fight) %>% 
              summarise(total_count2 = n(), .groups = 'drop') %>% 
              as.data.frame()

ggplot(data_grouped_5b, aes(fill=fight, y=total_count2, x=fight)) + 
    geom_bar(position="dodge", stat="identity") +
    ggtitle("Studying ") +
    facet_wrap(~race) +
    theme(legend.position="none") +
    xlab("")

```

Most respondents belonging to ethnic group 1 and 4 have not been in fights while most respondents from ethnic group 2 and 3have been in a fight. 


```{r}
#looking at age started drinking and if ever engaged in a physical fight 
ggplot(na.omit(data), aes(x=Age_Alc))+
  geom_bar(fill="forest green")+
  facet_wrap(~fight) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Engaging in a Physical Fight by Age when Individual Started Drinking") + xlab("Age")+ ylab("Frequency ")+ theme(plot.title = element_text(face="bold",hjust = 0.5))
```

The figure above shows that individuals who started drinking later in adolescence (around age 11 and 12) had never engaged in a physical fight. In comparison, individuals who started drinking earlier in adolescence (between ages 9 and 10) reported engaging in a physical fight at one point in their life. The overall trends show that more individuals who drank during adolescence have gotten into a physical fight. 

```{r}
#looking at age started drinking and if ever had suicidal thoughts
ggplot(na.omit(data), aes(x=Age_Alc))+
  geom_bar(fill="forest green")+
  facet_wrap(~suicide) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Thoughts of Suicide by Age when Individual Started Drinking") +  xlab("Age")+ ylab("Frequency") + theme(plot.title = element_text(face="bold",hjust = 0.5))
```

The results show that a majority of the youth have not had thoughts of committing suicide. The drinking age ranges for individuals who reported never having suicidal thoughts are more spread out with a majority reporting their first drinking experience being around ages 10 to 13. Contrastingly, individuals who reported having thoughts of suicide cluster around age 10 which may show a potential relationship between starting drinking earlier in adolescence and having future suicidal thoughts.  

```{r}
ttest2sample_alc = t.test(no_suicide$Age_Alc,yes_suicide$Age_Alc)
ttest2sample_alc

ttest2sample_alc1 = t.test(no_fight$Age_Alc,yes_fight$Age_Alc)
ttest2sample_alc1
```

The results of the two sample t-test looking at the difference in average ages between individuals who have and have not had thoughts about committing suicide show that there is a significant difference in the average ages of each group. The p-value = 0.00 which means that we can  accept the alternative hypothesis that there is a difference between suicidal and non-suicidal adolescents for the average age that they started drinking alcohol. The results also estimate the mean starting drinking age for non-suicidal individuals is about 11 years old while the estimated mean age for suicidal adolescents is around 10 years old when they started drinking. 

Similarly, the results of the two-sample t-test looking at the difference in average ages between individuals who have and have not gotten into a physical fight show that there is a significant difference in the average ages of each group. The p-value = 0.00 which means that we can accept the alternative hypothesis that there is a difference in average starting drinking ages between individuals who have and have not engaged in a physical fight. The resuls also estimate the mean starting drinking age for those who have engaged in a fight to be around 10 years old and around 11 years old for those who have not engaged in a physical fight. 

Overall, the results appear to show that the average starting age of drinking may be related to whether adolescents have had suicidal thoughts and have ever gotten into a physical fight. More specifically, it may be that the earlier adolescents start drinking the more likely they are to engage in these behaviors compared to adolescents who start drinking later in their life.  




```{r}
#looking at age started smoking marijuana and ever engage in a physical fight
ggplot(na.omit(data), aes(x=Age_Weed))+
  geom_bar(fill="forest green")+
  facet_wrap(~fight) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Engaging in a Physical Fight by Age Individual Started Smoking Marijuana") + xlab("Age")+ ylab("Frequency ")+ theme(plot.title = element_text(face="bold",hjust = 0.5))
```

The results of the figure above shows that individuals who have never gotten into a physical fight started smoking marijuana later in adolescence around age 11 and 12 while a majority of individuals who have gotten into a physical fight started smoking marijuana earlier in adolescence around ages 9 to 11. 

```{r}
#looking at age started smoking marijuana and suicidal thoughts
ggplot(na.omit(data), aes(x=Age_Weed))+
  geom_bar(fill="forest green")+
  facet_wrap(~suicide) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Thoughts of Suicide by Age Individual Started Smoking Marijuana") + xlab("Age")+ ylab("Frequency ")+ theme(plot.title = element_text(face="bold",hjust = 0.5))
```
The results show that a majority of individuals have not had suicidal thoughts and the ages they started smoking marijuana ranges between 10 and 13. Interestingly, for the small majority of individuals who did report having thoughts of commiting suicide started smoking marijuana around age 9. This may reveal a potential relationship between smoking marijuana really early in adolescence and the consequences it may have on their mental health in terms of having suicidal thoughts. 

```{r }
ttest2sample_weed = t.test(no_suicide$Age_Weed,yes_suicide$Age_Weed)
ttest2sample_weed

ttest2sample_weed1 = t.test(no_fight$Age_Weed,yes_fight$Age_Weed)
ttest2sample_weed1
```

The results of the two sample t-test looking at the difference in average ages between individuals who have and have not had thoughts about committing suicide show that there is a significant difference in the average ages of each group. The p-value = 0.00 which means that we can  accept the alternative hypothesis that there is a difference between suicidal and non-suicidal adolescents for the average age that they started smoking marijuana. The results also estimate the mean starting smoking age for non-suicidal individuals is about 11 years old while the estimated mean age for suicidal adolescents is around 10 years old for when they started smoking marijuana. 

Similarly, the results of the two-sample t-test looking at the difference in average ages between individuals who have and have not gotten into a physical fight show that there is a significant difference in the average starting smoking ages of each group. The p-value = 0.00 which means that we can accept the alternative hypothesis that there is a difference in average starting smoking ages between individuals who have and have not engaged in a physical fight. The results also estimate the mean starting smoking age for those who have engaged in a fight to be around 10 years old and around 11 years old for those who have not engaged in a physical fight. 

Overall, the results appear to show that the average starting age of smoking marijuana may be related to whether adolescents have had suicidal thoughts and/or have ever gotten into a physical fight. More specifically, it may be that the earlier adolescents start marijuana the more likely they may be to engage in these behaviors compared to adolescents who start smoking marijuana later in their life.  

```{r}
#looking at age started drinking and if ever engaged in a physical fight 
ggplot(na.omit(data), aes(x=age_sex))+
  geom_bar(fill="forest green")+
  facet_wrap(~fight) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Engaging in a Physical Fight by Age when Individual Started Having Romantic Relationships") + xlab("Age")+ ylab("Frequency ")+ theme(plot.title = element_text(face="bold",hjust = 0.5))
```

The results show that individuals who have not engaged in a physical fight reported having romantic relationships between ages 12 and 13 while individuals who have engaged in a physical fight reported having romantic relationships around 10 and 11 years old. These results may show that the later an individual starts having romantic relationships, the less likely they may be to get into physical fights and vice versa. 

```{r}
#looking at age started drinking and if ever had suicidal thoughts
ggplot(na.omit(data), aes(x=age_sex))+
  geom_bar(fill="forest green")+
  facet_wrap(~suicide)  + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Thoughts of Suicide by Age when Individual Started Having Sex") + xlab("Age")+ ylab("Frequency ")+ theme(plot.title = element_text(face="bold",hjust = 0.5))
```

The results above show that a majority of individuals who did not report having suicidal thoughts started having romantic relationships between ages 11 and 13 while a majority of individuals who did report having suicidal thoughts started having romantic relationships around age 10. The results may provide insight into how engaging in romantic relationships earlier in adolescence may lead to future suicidal thoughts for youth. 


```{r}
ttest2sample_sex = t.test(no_suicide$age_sex,yes_suicide$age_sex)
ttest2sample_sex

ttest2sample_sex1 = t.test(no_fight$age_sex,yes_fight$age_sex)
ttest2sample_sex1
```


The results of the two sample t-test looking at the difference in average ages between individuals who have and have not had thoughts about committing suicide show that there is a significant difference in the average ages of each group. The p-value = 0.00 which means that we can  accept the alternative hypothesis that there is a difference between suicidal and non-suicidal adolescents for the average age that they started having romantic relationships. The results also estimate the mean age for first romantic relationships for non-suicidal individuals is about 11 years old while the estimated mean age for suicidal adolescents is around 10 years old for when they started having romantic interests.

Similarly, the results of the two-sample t-test looking at the difference in average ages between individuals who have and have not gotten into a physical fight show that there is a significant difference in the average ages of when they had their first romantic interests. The p-value = 0.00 which means that we can accept the alternative hypothesis that there is a difference in average starting smoking ages between individuals who have and have not engaged in a physical fight. 

Overall, the results appear to show that the average starting age of romantic relationships may be related to whether adolescents have suicidal thoughts and/or have ever gotten into a physical fight. More specifically, it may be that the earlier adolescents start having romantic relationships the more likely they may be to engage in these behaviors compared to adolescents who have romantic interests later in their life.

```{r}
race_vs_multipartners <- aov(multiple_partners ~ race, data =data )
xkabledply(race_vs_multipartners)
```
Since the p-value of 0.000 is less than the significance level of 0.05, we rejects the null hypothesis that there is NO difference in the average number of romantic partners an individual has had between races.

```{r}
race_vs_physicalactivity <- aov(physical_activity ~ race, data =data )
xkabledply(race_vs_physicalactivity)
```
Since the p-value of 0.000 is less than the significance level of 0.05, we rejects the null hypothesis that there is NO difference in the average days of physical activity between races.

```{r}
race_vs_HrsOfSleep <- aov(Hrs_Sleep ~ race, data = data)
xkabledply(race_vs_HrsOfSleep)
```
Since the p-value of 0.000 is less than the significance level of 0.05, we rejects the null hypothesis that there is NO difference in the average hour of sleeps of different individuals between races.

```{r}
data$race <- recode(data$race, '1' = 'White', '2' = 'Black or African American','3'='Hispanic/Latino','4'='All Other Races')
```

```{r}
agg_df <- aggregate(cbind(data$Hrs_Sleep,data$physical_activity,data$multiple_partner,data$Age_Weed,data$age_sex,data$Age_Alc) ~ data$race, FUN=mean)

colnames(agg_df) = c('Race','Hours of Sleep','Days of Physical Activity','Sexual Partners','Age Smoking Marijuana','Age Having Romantic Relations','Age Drinking Alcohol')

customGreen0 = "#DeF7E9"
customGreen = "#71CA97"
formattable(agg_df, align =c("l","c","c","c","c", "c", "r"), list(
  'Race' = formatter("span", style = ~ style(color = "dark gray",font.weight = "bold")), 
  'Hours of Sleep' = color_tile(customGreen,customGreen0),
  'Days of Physical Activity'= color_tile(customGreen,customGreen0),
  'Sexual Partners'= color_tile(customGreen,customGreen0),
  'Age Smoking Marijuana'= color_tile(customGreen,customGreen0),
  'Age Having Romantic Relations'= color_tile(customGreen,customGreen0),
  'Age Drinking Alcohol'= color_tile(customGreen,customGreen0)
))
```
The average  hour of sleep by white race is 6.32 hours which is highest among the all the races whereas the average hour of sleep by Hispanic/Latino and Black or African American are 6.26 and 5.516 hours. All other Races sleep 6.11 average hour of sleep.

The average days of physical activity between white races is 3.87 whereas the average days of physical activity between Hispanic/Latino, Black or African American and All Other Races 
are 5.776, 6.032 and 3.90.

The average number of romantic partners by white races has had is 2.83 whereas the average numbers of romantic partner by Hispanic/Latino, Black or African American and All Other Races has had are 2.64, 5.233 and 2.707 respectively.

The average age of white people when started smoking marijuana is 11.50 whereas the average age of Hispanic/Latino, Black or African American and All Other Races when they started smoking marijuana are 11.51, 9.68 and 11.51 respectively.

The average age of white people when having sex is 11.41 whereas the average age of Hispanic/Latino, Black or African American and All Other Races when they started having sex are 11.60,10.30 and 10.95 respectively.

The average age of white people when they started drinking alcohol is 11.11 whereas the average age of Hispanic/Latino, Black or African American and All Other Races  when they started drinking alcohol are 11.15,10.09 and 11.28 respectively.

